---
title: "Rendering Pandoc Grid Tables with GridTable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Pandoc Grid Tables with GridTable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment  = ""      # show the ASCII tables verbatim, no "## " prefix
)
library(GridTable)
```

## What `GridTable` does

`GridTable` turns a `data.frame`, `data.table`, or `knitr::kable` into a
plain-text **pandoc grid table** — the ASCII boxes drawn with `+ | - = :` that
pandoc understands as a real table:

```{r basic}
df <- data.frame(name = c("Alice", "Bob"), score = c(91.5, 88))
print(GridTable(df))
```

A few things already happened automatically:

* the **column names** became the first row, and a header rule (`=`) was drawn
  under them;
* the numeric `score` column was formatted to a fixed number of decimals;
* every column got a sensible width.

The engine measures text with `nchar(type = "width")`, so wide (CJK) characters
count as two columns and stay aligned — more on that at the end.

A `GridTable` is just a `data.table` carrying its render state as attributes.
You build it once, tweak it by reference, and `print()` it (or call
`toString()` to get the lines as a character vector).

## Per-column alignment

Pass `align` as one code per column (`l`eft, `r`ight, `c`entre), or a single
code recycled to all columns. Numeric columns default to right, character
columns to left. The alignment governs the **body** text as well as the header
rule's colons:

```{r align}
df <- data.frame(item = c("apples", "pie"), qty = c(4, 12), note = c("fresh", "warm"))
print(GridTable(df, align = "lcr"))
```

## Header and footer rules

`header` is the row index the header rule is drawn under (the default prepends
the column names and uses `header = 1`). `footer` draws a rule above the given
row and along the table bottom — handy for a totals row:

```{r footer}
sales <- data.frame(
  region = c("North", "South", "Total"),
  amount = c(120, 95, 215)
)
print(GridTable(sales, footer = 3))
```

## Captions

Anything passed through `...` is stored as an attribute; `caption` is rendered
above the table:

```{r caption}
print(GridTable(df, caption = "Table: today's order"))
```

## Merging cells

`merge_cells(tbl, i, j)` registers a rectangular merged region spanning rows `i`
and columns `j`. It modifies the table **by reference**. Remember that — with
the default auto-header — the first *data* row is row 2.

### Column span (a header across columns)

```{r colspan}
m <- matrix(
  c("Location",   "Temp", "Temp",
    "city",       "min",  "max",
    "Antarctica", "-89.2", "19.8"),
  ncol = 3, byrow = TRUE
)
gt <- GridTable(m, header = 2, align = "lrr")
merge_cells(gt, i = 1, j = 2:3)   # "Temp" spans columns 2-3
print(gt)
```

### Row span

```{r rowspan}
gt <- GridTable(data.frame(group = c("A", "A", "B"), value = c(1, 2, 3)))
merge_cells(gt, i = 2:3, j = 1)   # merge the first data column across rows 2-3
print(gt)
```

### Row and column span at once

```{r rowcolspan}
m <- matrix(c("H1",  "H2",  "H3",
              "BIG", "BIG", "x",
              "BIG", "BIG", "y"),
            ncol = 3, byrow = TRUE)
gt <- GridTable(m, header = 1)
merge_cells(gt, i = 2:3, j = 1:2)
print(gt)
```

## Merge options

`merge_cells()` takes three options that change how a region's content is laid
out.

### `drop_content` — keep only the anchor cell

By default the members of a region are joined together. With
`drop_content = TRUE` only the top-left (anchor) cell's text is kept:

```{r drop-content}
m <- matrix(c("H1", "H2", "A", "B", "C", "D"), ncol = 2, byrow = TRUE)
gt <- GridTable(m, header = 1)
merge_cells(gt, i = 2:3, j = 1, drop_content = TRUE)
print(gt)
```

### `middle` — vertical centring

```{r middle}
gt <- GridTable(data.frame(group = c("x", "y", "z"), value = c(1, 2, 3)))
merge_cells(gt, i = 2:3, j = 1, middle = TRUE)
print(gt)
```

### `wrap` — pandoc continuation lines

`wrap` appends pandoc's continuation backslash to every line but the last, so a
multi-line cell renders as one logical line when the table is re-processed by
pandoc:

```{r wrap}
m <- matrix(c("H1",                             "H2",
              "first line\nsecond line\nthird", "y"),
            ncol = 2, byrow = TRUE)
gt <- GridTable(m, header = 1)
merge_cells(gt, i = 2, j = 1, wrap = TRUE)
print(gt)
```

To list the regions registered on a table, call `merge_cells(tbl)` with no
`i`/`j`; to remove one, pass `cancel = TRUE` (with the same `i`/`j`) or the
region's name.

## Adjusting widths and heights: the `set_attr` DSL

The engine sizes everything to fit, but you can pin or nudge a column width or
row height with `set_attr()`. Besides a plain numeric vector, `width` and
`height` accept an adjustment string `<index><op><operand>`, where the index is
a number or a letter (`A` = 1, `B` = 2, …) and the op is one of `+ - * / =`:

```{r set-attr}
gt <- GridTable(data.frame(a = c("x", "y"), b = c(1, 2)))
set_attr(gt, "width", "1+3")   # widen column 1 by three display columns
print(gt)
```

`set_attr()` can set any attribute, and several at once through `...`
(e.g. `set_attr(gt, align = "c", caption = "...")`).

## Starting from a `knitr::kable`

If you already have a pipe- or simple-format `kable`, `kable_to_grid()`
reverse-parses it (recovering the column boundaries from the separator line)
into a `GridTable` you can then merge, align, or caption:

```{r kable}
kbl <- knitr::kable(
  data.frame(City = c("Beijing", "Xian"), Pop = c(2189, 1295)),
  format = "pipe"
)
print(kable_to_grid(kbl))
```

## CJK width awareness

Because widths are measured in display columns, tables mixing wide and narrow
characters stay aligned:

```{r cjk}
df <- data.frame(city = c("北京", "上海", "x"), pop = c(2189, 2487, 5))
print(GridTable(df, align = "lr"))
```

## Summary

The public surface is small:

* `GridTable()` / `kable_to_grid()` — build a table;
* `merge_cells()` — merge regions (`drop_content` / `middle` / `wrap`);
* `set_attr()` — pin or adjust widths, heights, alignment, captions;
* `print()` / `toString()` — render it.

Everything is computed up front and laid out with local, CJK-aware decisions, so
even tables that combine row+column spans with footers and wrapping render
cleanly.
